---
title: "Cancer_first_ICD10"
author: "Yuchen Zhang"
date: "1/26/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(tidyverse)
library(data.table)
library(arsenal)
```

## Read in data

```{r}
# Change the local drive to the cluster drive
# cluster location /scratch/yz92460/ukb/omega_data/raw.csv
# local location /Users/yuchen/Desktop/UKBiobank/Cancer/Data/
ID_new = omega_new$ID

#cancer_new = fread("/Users/yuchen/Desktop/Cancer/Data/cancer_new.csv", header=TRUE, sep=",")
health_new = fread("/Users/yuchen/Desktop/Cancer/Data/health_new2.csv", header=TRUE, sep=",")

#omega <- fread("/Users/yuchen/Desktop/Cancer/Data/omega.csv", header=TRUE, sep=",") # need new
omega_new = fread("/Users/yuchen/Desktop/Cancer/Data/omega_new.csv", header=TRUE, sep=",")

#death <- fread("/Users/yuchen/Desktop/Cancer/Data/death.csv", header=TRUE, sep=",")
death_new <- fread("/Users/yuchen/Desktop/Cancer/Data/death_new2.csv", header=TRUE, sep=",")

fish <- fread("/Users/yuchen/Desktop/UKBiobank/Data/data_fish_0210.csv", header=TRUE, sep=",")
fish = fish %>% select(ID, Fish_oil_baseline)

covar <- fread("/Users/yuchen/Desktop/Cancer/Data/covar.csv", header=TRUE, sep=",")
covar = covar %>% filter(ID %in% ID_new)
covar = covar %>% select(-Fish_oil_baseline, -omega3_pct, -omega6_pct, -omega_ratio2)

loss = fread("/Users/yuchen/Desktop/Cancer/Data/loss.csv", header=TRUE, sep=",")
loss = loss %>% filter(ID %in% ID_new)

others = fread("/Users/yuchen/Desktop/Cancer/Data/others.csv", header=TRUE, sep=",")
```


Merge the dataset

```{r}
mydata = omega_new %>% left_join(covar, by = "ID")
mydata = mydata %>% left_join(death_new, by = "ID")
mydata = mydata %>% left_join(fish, by = "ID")
mydata = mydata %>% left_join(loss, by = "ID")
mydata = mydata %>% left_join(others, by = "ID")
```


date

```{r}
library(stringr)
# Select date
cancer_base_wide_date = health_new %>% select(ID, starts_with('cancer_date'))
# change wide format to long format
cancer_base_long_date = melt(setDT(cancer_base_wide_date), measure.vars=list(c(2:23)), variable.name='instance', value.name=c('date'))[order(ID)]

cancer_base_long_date$instance = sub("cancer_date", "", cancer_base_long_date$instance)
cancer_base_long_date$instance = str_extract(cancer_base_long_date$instance, "[0-9]+")
```


```{r}
# Select ICD-10
cancer_base_wide_10 = health_new %>% select(ID, starts_with('cancer_type_10'))
# change wide format to long format
cancer_base_long_10 = melt(setDT(cancer_base_wide_10), measure.vars=list(c(2:23)), variable.name='instance', value.name=c('type_10'))[order(ID)]

cancer_base_long_10$instance = sub("cancer_type_10icd", "", cancer_base_long_10$instance)
cancer_base_long_10$instance = str_extract(cancer_base_long_10$instance, "[0-9]+")
```


check with icd-9

```{r}
# Select ICD-9
cancer_base_wide_9 = health_new %>% select(ID, starts_with('cancer_type_9'))
# change wide format to long format
cancer_base_long_9 = melt(setDT(cancer_base_wide_9), measure.vars=list(c(2:16)), variable.name='instance', value.name=c('type_9'))[order(ID)]

cancer_base_long_9$instance = sub("cancer_type_9icd", "", cancer_base_long_9$instance)
cancer_base_long_9$instance = str_extract(cancer_base_long_9$instance, "[0-9]+")
```

combine date and icd-10

```{r}
cancer = cancer_base_long_date %>% left_join(cancer_base_long_10, by = c("ID", "instance"))
cancer = cancer %>% left_join(cancer_base_long_9, by = c("ID", "instance"))
```

select the first 15 instances
```{r}
cancer_15 = cancer %>% group_by(ID) %>% slice(c(1:15)) 
```

map with icd-9
```{r}
check = cancer_15
or = check %>% filter(is.na(type_9)==FALSE | is.na(type_10)==FALSE) # 167947


only_10 = check %>% filter(is.na(type_9)==TRUE & is.na(type_10)==FALSE & is.na(date)==FALSE) # 152700
only_9 = check %>% filter(is.na(type_10)==TRUE & is.na(type_9)==FALSE & is.na(date)==FALSE) # 15247
both_have = check %>% filter(is.na(type_9)==FALSE & is.na(type_10)==FALSE) # 0

```

check 16-22
```{r}
cancer_remain = cancer %>% group_by(ID) %>% slice(c(16:22))
missing = cancer_remain %>% filter(is.na(type_10)==FALSE | is.na(date)==FALSE) # 10
```

begin select the first occurrence
```{r}
cancer_1 = cancer %>% arrange(ID, date) %>% filter(is.na(date)==FALSE)
cancer_1 = cancer_1 %>% group_by(ID) %>% slice(1)
cancer_1 = cancer_1 %>% rename(cancer_date_1 = date)
# create the event variable for those who had cancer
cancer_1 = cancer_1 %>% mutate(status = 1) # 123776

#### Deal with C44
cancer_1$icd10_ini = substr(cancer_1$type_10,1,1)
cancer_1$icd10_num = substring(cancer_1$type_10,2,3)
cancer_1$icd10_num = as.numeric(cancer_1$icd10_num)
cancer_1_C44 = cancer_1 %>% filter(icd10_num == 44) # 31181
cancer_1 = cancer_1 %>% filter(!ID %in% cancer_1_C44$ID) # 92595
```


merge the cancer data and covar

```{r}
cancer_final = mydata %>% left_join(cancer_1, by = "ID")
```


Create the table with only first occurrence data

```{r}
# Exclude those had cancer at baseline
cancer_final$base_date = as.Date(cancer_final$base_date, format = "%Y%m%d", origin=as.Date("1970-01-01"))
cancer_have = cancer_final %>% filter(is.na(cancer_date_1)==FALSE)
cancer_final2 = cancer_final  %>% filter(base_date <= cancer_date_1)

# number of patients had cancer 92595
num_pre = nrow(cancer_have)
# number of patients not had cancer at baseline 54861
num_post = nrow(cancer_final2) 
# ID need to be exclude 37734
ID_exc = cancer_have %>% select(ID) %>% filter(!ID %in% cancer_final2$ID)
# number of patients had cancer at baseline 37734
num_have = nrow(ID_exc)

print(c(num_pre, num_post, num_have))
```

```{r}
mydata_cancer = cancer_final %>% filter(!ID %in% ID_exc$ID) %>% mutate(status = ifelse(is.na(status)==TRUE, 0, 1))
# those include does not have icd-9 
```


### Exclude loss to follow up

```{r}
mydata_cancer2 = mydata_cancer %>% filter(is.na(Lost_follow_up) == "TRUE")
# print the numbers of tuples before and after exclude loss to follow up
num_raw = nrow(mydata_cancer)
num_after_loss = nrow(mydata_cancer2) #-1227
print(c(num_raw, num_after_loss))
```

Total cancer
```{r}
sum(mydata_cancer2$status)
```


Merge the dataset with covariates

```{r}
# create the ethnic group
mydata_cancer = mydata_cancer2 %>% 
    mutate(eth_group = ifelse(ethnicity %in% c(1,1001,2001,3001,4001),"White",
                              ifelse(ethnicity %in% c(2,1002,2002,3002,4002),"Mixed",
                                    ifelse(ethnicity %in% c(3,1003,2003,3003,4003),"Asian or Asian British",
                                          ifelse(ethnicity %in% c(4,2004,3004),"Black or Black British",
                                                ifelse(ethnicity %in% c(5),"Chinese",
                                                    ifelse(ethnicity %in% c(6),"Other ethnic group",
                                                                            "DK/NR")))))))
# Combine the ethnicity
mydata_cancer = mydata_cancer %>% 
    mutate(eth_group_new = ifelse(eth_group == "White","White",
                                  ifelse(eth_group == "Black or Black British","Black",
                                         ifelse(eth_group %in% c("Asian or Asian British", "Chinese"),"Asian",
                                                ifelse(eth_group %in% c("Mixed", "Other ethnic group"),"Others",
                                                       NA)))))  %>% select(-c(ethnicity, eth_group))
mydata_cancer$Alcohol_status[mydata_cancer$Alcohol_status==-3] = NA
mydata_cancer$Smoking_status[mydata_cancer$Smoking_status==-3] = NA



# Calculate the time interval
mydata_cancer = mydata_cancer %>% mutate(cancer_date_1 = ifelse(is.na(cancer_date_1)==TRUE, as.Date("2022-12-19"), cancer_date_1))
mydata_cancer = mydata_cancer %>% mutate(date_of_death = ifelse(is.na(date_of_death)==TRUE, as.Date("2022-12-19"), date_of_death))
mydata_cancer$cancer_date_1 = as.Date(mydata_cancer$cancer_date_1, format = "%Y%m%d", origin=as.Date("1970-01-01"))
mydata_cancer$date_of_death = as.Date(mydata_cancer$date_of_death, format = "%Y%m%d", origin=as.Date("1970-01-01"))
# end of study (which comes first)
mydata_cancer = mydata_cancer %>% mutate(end_date = ifelse(date_of_death < cancer_date_1, date_of_death, cancer_date_1))
mydata_cancer$end_date = as.Date(mydata_cancer$end_date, format = "%Y%m%d", origin=as.Date("1970-01-01"))
mydata_cancer$time_to_cancer = difftime(mydata_cancer$end_date, mydata_cancer$base_date, units = "days") # in days
mydata_cancer$time_to_cancer_year = round(mydata_cancer$time_to_cancer/365.2425,2)
mydata_cancer$time_to_cancer_year = as.numeric(mydata_cancer$time_to_cancer_year)
# cancer subgroup
mydata_cancer = mydata_cancer %>% 
    mutate(cancer_sub = 
      ifelse(icd10_num >= 0 & icd10_num <= 14 ,"01_head",
         ifelse(icd10_num == 15, "02_esopha",
            ifelse(icd10_num == 16,"03_stomach",
               ifelse(icd10_num == 18,"04_colon",
                  ifelse(icd10_num == 19 | icd10_num == 20 ,"05_rectum",
                    ifelse(icd10_num >= 22 & icd10_num <= 24,"06_hepatob",
                     ifelse(icd10_num == 25,"07_pancrea",
                      ifelse(icd10_num == 33 | icd10_num == 34,"08_lung",
                        ifelse(icd10_num == 43,"09_mal_melan",
                          ifelse(icd10_num ==49,"10_connect",
                            ifelse(icd10_num == 50,"11_breast",
                              ifelse(icd10_num == 54 | icd10_num == 55,"12_uterus",
                               ifelse(icd10_num == 56,"13_ovary",
                                 ifelse(icd10_num == 61,"14_prostate",
                                   ifelse(icd10_num == 64 | icd10_num == 65,"15_kidney",
                                     ifelse(icd10_num == 66 | icd10_num == 67,"16_bladder",
                                       ifelse(icd10_num >= 70 & icd10_num <= 72,"17_brain",
                                         ifelse(icd10_num == 73,"18_thyroid",
                                           ifelse(icd10_num >= 81 & icd10_num <= 96,"19_lymphoma",NA
                                                                 ))))))))))))))))))))
mydata_cancer = mydata_cancer %>% filter(is.na(time_to_cancer_year) == FALSE)
# Save the data
# cluster location /scratch/yz92460/ukb/omega_data/raw.csv
# local location /Users/yuchen/Desktop/UKBiobank/Cancer/Data/
write.csv(mydata_cancer,file="/Users/yuchen/Desktop/Cancer/Data/mydata_cancer.csv",row.names = F)

```

Get the min, max, median, mean for individuals developing cancer

```{r}
test_c = mydata_cancer %>% filter(status == 1)
print(median(mydata_cancer$time_to_cancer_year))
```

```{r}
hist(test_c$time_to_cancer_year, xlab = "Year", main = "Distribution of time to first cancer")
```


### Get the data with complete information on plasma PUFAs

```{r}
plasma_cancer = mydata_cancer %>% filter(is.na(omega_ratio) == FALSE) # 253138
plasma_excnum = nrow(mydata_cancer) - nrow(plasma_cancer)
print(plasma_excnum)
# Save the data
# cluster location /scratch/yz92460/ukb/omega_data/raw.csv
# local location /Users/yuchen/Desktop/UKBiobank/Cancer/Data/
write.csv(plasma_cancer,file="/Users/yuchen/Desktop/Cancer/Data/plasma_cancer.csv",row.names = F)
```

Baseline characteristics table for plasma data

```{r results="asis"}
data_add = plasma_cancer %>% mutate(omega3_pctg = ntile(omega3_pct, 5))
data_add = data_add %>% mutate(omega6_pctg = ntile(omega6_pct, 5))
data_add = data_add %>% mutate(omega_ratiog = ntile(omega_ratio, 5))


data_add$sex = as.factor(data_add$sex)
data_add$Fish_oil_baseline = as.factor(data_add$Fish_oil_baseline)
data_add$Alcohol_status = as.factor(data_add$Alcohol_status)
data_add$Smoking_status =as.factor(data_add$Smoking_status)
data_add$IPAQ_activity = as.factor(data_add$IPAQ_activity)
data_add$omega_ratiog = as.factor(data_add$omega_ratiog)

plasma_selected = data_add[, c(3,4,25,5,7,9,8,10,11,12,13,32)]
demo_plasma = tableby( omega_ratiog~., data = plasma_selected, control=tableby.control(numeric.test="anova", cat.test="chisq"))

summary(demo_plasma, pfootnote = T, digits = 1)
```

### Get the data with complete information on fish oil supplementation

```{r}
fish_cancer = mydata_cancer %>% filter(is.na(Fish_oil_baseline) == FALSE) # 457683
fish_excnum = nrow(mydata_cancer) - nrow(fish_cancer)
print(fish_excnum)
# Save the data
# cluster location /scratch/yz92460/ukb/omega_data/raw.csv
# local location /Users/yuchen/Desktop/UKBiobank/Cancer/Data/
write.csv(fish_cancer,file="/Users/yuchen/Desktop/Cancer/Data/fish_cancer.csv",row.names = F)
```

Baseline characteristics table for fish oil supplementation data

```{r results="asis"}
fish_cancer$sex = as.factor(fish_cancer$sex)
fish_cancer$Fish_oil_baseline = as.factor(fish_cancer$Fish_oil_baseline)
fish_cancer$Alcohol_status = as.factor(fish_cancer$Alcohol_status)
fish_cancer$Smoking_status =as.factor(fish_cancer$Smoking_status)

fish_selected = fish_cancer[, c(11,12,13,14)]
demo_fish = tableby(Fish_oil_baseline~ ., data = fish_selected, control=tableby.control(numeric.test="anova", cat.test="chisq"))

summary(demo_fish, pfootnote = T, digits = 1)
```

